---
title: "Spatial Analysis and Mapping Workshop"
author: "Zack W Almquist (University of Minnesota)"
date: "November 9, 2016"
output: beamer_presentation
---

```{r,include=FALSE}
require(knitr)
opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE,tidy=TRUE)
options(width=60)
```

## R spatial data handling and GIS activities

\begin{itemize}
\item Roger Bivand's book \emph{Applied Spatial Data Analysis with R} 
\item Contains tools for handling many (most?) different spatial data formats %ESRI shapefiles 
\item Contains tools for managing standard GIS activities such as plotting and overlays
\item Contains tools for for statistical spatial analysis
\item Contains tools for reading shape and raster files
\end{itemize}

## The UScensus2000-suite of Packages

\setbeamercovered{transparent}
\begin{block}{UScensus packages in R}
\begin{itemize}
\item<1-> 13 packages
\begin{itemize}
\item UScensus1990\{blkgrp\}
\item UScensus2000\{block, blockgroup, tract, cdp\}
\item UScensus2010\{block, blockgroup, tract, county, cdp\}
\end{itemize}
\item<2-> 3 packages of helper functions
\item<3-> 10 packages of polygon/shapefiles and demographic data
\item<4-> All data from US Census Bureau's \alert{SF1} files and \alert{TigerLine Shapefiles}
\end{itemize}
\end{block}

## Structure of the UScensus2000 Packages

\begin{block}{}
\tiny{
\setlength{\unitlength}{.7cm}
\begin{picture}(6,6)(-3,.5)

\put(0,5){\framebox(4,1)[0,0]{UScensus2000}}
\put(5,5){\framebox(4,1)[0,0]{UScensus2000add}}


\put(2,4){\line(0,1){1}}
\put(7,4){\line(0,1){1}}
\put(-1,4){\line(1,0){11}}

\put(-1,4){\vector(0,-1){1}}
\put(2,4){\vector(0,-1){1}}
\put(7,4){\vector(0,-1){1}}
\put(10,4){\vector(0,-1){1}}

\put(-2.2,1.9){\framebox(3.45,1)[0,0]{UScensus2000blk}}
\put(1.4,1.9){\framebox(3.45,1)[0,0]{UScensus2000blkgrp}}
\put(5,1.9){\framebox(3.45,1)[0,0]{UScensus2000tract}}
\put(8.6,1.9){\framebox(3.45,1)[0,0]{UScensus2000cdp}}
\end{picture}
}
\end{block}

## Organization of the US Census
\begin{block}{}
\setlength{\unitlength}{.8cm}
\begin{picture}(3,7)(-2,.2)

\put(-1,6){\framebox(12,1)[0,0]{County}}
\put(0,4){\framebox(10,1)[0,0]{Tract}}
\put(1,2){\framebox(8,1)[0,0]{Block Group}}
\put(3,0){\framebox(4,1)[0,0]{Block}}

\put(5,1){\vector(0,1){1}}
\put(5,3){\vector(0,1){1}}
\put(5,5){\vector(0,1){1}}

\end{picture}

\vspace{.1in}

\end{block}

## Organization of the US Census

\begin{center}
\includegraphics[width=.7\linewidth]{figures/geography}
\end{center}

## Available Data

\begin{block}{Via The Comprehensive R Archive Network
(CRAN)\\ \url{http://cran.r-project.org/}}
\begin{itemize}
\item Tract (UScensus2000tract)
\item Census Designated Place (UScensus2000cdp)
\item UScensus2010
\end{itemize}
\end{block}

## Available Data

\begin{block}{Via R-Forge\\
\url{https://r-forge.r-project.org/R/?group_id=2022}}
\begin{itemize}
\item Block, Block Group, Tract, County, CDP (UScensus2000 and UScensus2010)
\end{itemize}
\end{block}

## Available Data
\begin{block}{Via NCASD Lab\\ \url{http://lakshmi.calit2.uci.edu/census2000/}}
\begin{itemize}
\item Block (UScensus1990blkgrp)
\item Block  (UScensus2000blk and UScensus2010blk)
\end{itemize}
\end{block}

## Installation

\tiny
```{r, eval=FALSE}
install.packages("UScensus2010", 
                 repos="http://R-Forge.R-project.org")
install.packages("UScensus2010county", 
                 repos="http://R-Forge.R-project.org")
install.packages("UScensus2010tract", 
                 repos="http://R-Forge.R-project.org")
install.packages("UScensus2010cdp", 
                 repos="http://R-Forge.R-project.org")
```

## Structure of the UScensus2000 Data-Packages

\begin{block}{}
 \setlength{\unitlength}{.8cm}
\begin{picture}(6,6)(-1,.2)
\put(1.9,5){\framebox(7.6,1)[0,0]{Package (e.g., UScensus2000tract)}}
\put(1.9,3){\framebox(7.5,1)[0,0]{State (e.g., california.tract)}}
\put(-.5,.4){\framebox(13,1.5)[0,0]{data and polygons}}
\put(0,.6){(e.g., california.tract@data or california.tract@polygons)}

\put(5.5,3){\vector(0,-1){1}}
\put(5.5,5){\vector(0,-1){1}}
\end{picture}
\end{block}
\vspace{-.2in}
\begin{block}{}
\begin{itemize}
\item All data is stored as \alert{SpatialPolygonsDataframe} object
\item \alert{data} is a data.frame object with {\bf ID} (factors) and {\bf demographic} (numeric) values 
\item \alert{polygons} is a list of the spatial data
\end{itemize}
\end{block}

## Loading the Data
\tiny
```{r,cache=FALSE}
 library(UScensus2000)
 data(california.tract)
 summary(as(california.tract,"SpatialPolygons"))
```

## Loading the Data
\tiny
```{r}
 names(california.tract)
```

## Help

```{r,eval=FALSE}
help(california.tract)
```
## Help

\includegraphics[width=.65\linewidth]{figures/rhelpss}

## Examples

\begin{center}
Useful Functions in the UScensus2000 Package
\end{center}

## UScensus2000

\begin{block}{Functions}
\begin{center}
\begin{itemize}
\item choropleth()
\item county()
\item MSA()
\item city()
\item poly.clipper()
\item demographics()
\end{itemize}
\end{center}
\end{block}

## Choropleth Maps
\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center',cache=FALSE}
UScensus2000::choropleth(california.tract, main="2000 US Census Tracts \n California", border="transparent")
```

## county

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center'}
la.county <- county(name="los angeles",
                      state="ca", level="tract")
plot(la.county)
 ```
 
 ## MSA

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center'}
losangeles.msa<-MSA(msaname="Los Angeles",
                    state="CA",level="tract")
plot(losangeles.msa,border=rgb(0,0,0,.4))
 ```
 
## City
 
 \tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center'}
 losangeles<-city(name="los angeles",
                    state="ca")
plot(losangeles)
 ```
 
 ## demographics
 
 \tiny
```{r}
laMSAarea<-demographics(dem=c("pop2000","white","black"),
                        "CA",level="msa",msaname="Los Angeles")
laMSAarea
 ```
 
 ## demographics
 \tiny
 ```{r}
ca.cdp<-demographics(dem=c("pop2000","white","black",
        "hh.units","hh.vacant"), "CA",level="cdp") 
 ##Alphabetic order the first 10 CDPs 
ca.cdp[order(rownames(ca.cdp))[1:10],]
```

## rgdal

\begin{itemize}
\item Requires GDAL and proj4 packages (these must be installed at the OS level)
\item A it of a pain, but worth it!
\item rgdal allows one to read in ESRI shape files
\item rgdal allows one to re-project spatial data 
\item<2> Let's do an example!
\end{itemize}

## rgdal

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center',cache=FALSE}
library(rgdal)
UScounty<-readOGR(dsn="data/co99_d00_shp",layer="co99_d00",stringsAsFactors=FALSE)
proj4string(UScounty)<-" +proj=longlat +ellps=GRS80 +datum=NAD83 +no_defs +towgs84=0,0,0"
```
 

## rgdal

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center'}
plot(UScounty)
```
 
## rgdal: Translate to an equal area projection

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center'}
UScounty<-spTransform(UScounty,CRS(
" +proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 
+towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))
```

## rgdal

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center',cache=FALSE}
library(UScensus2010)
data(countyfips)
UScounty$fips<-paste(UScounty$STATE,UScounty$COUNTY,sep="")
ah.fips<-unique(substr(countyfips$fips[countyfips$statename%in%c("hawaii","alaska")],1,2))
conUS<-UScounty[!UScounty$STATE%in%ah.fips,]
alaska<-UScounty[UScounty$STATE==ah.fips[1],]
hawaii<-UScounty[UScounty$STATE==ah.fips[2],]
```

## rgdal

\tiny
```{r,fig.width = 5, fig.height = 5,fig.align='center'}
par(mar=c(1, 1, 0, 0) + 0.1,mgp=c(0, 0.2,0),tcl=-.2)
plot(conUS,border=rgb(0,0,0,.2),col=rgb(0,0,0,.2))
par(fig=c(.25, 2, .2, 2.1)/10,new=TRUE,mgp=c(0,0.2,0),tcl=-.2)
plot(alaska,border=rgb(0,0,0,.3),col=rgb(0,0,0,.2))
box()
par(fig=c(2.25, 4, .2, 2.1)/10,new=TRUE,mgp=c(0,0.2,0),tcl=-.2)
plot(hawaii,border=rgb(0,0,0,.3),col=rgb(0,0,0,.05))
box()
```

## US Census API

\begin{itemize}
\item Access to US Demographics SF1, SF3 (1990, 2000) and ACS (2010)
\item Much More Too!!
\item \url{http://www.census.gov/data/developers/data-sets.html}
\item Requires User Key (\url{http://api.census.gov/data/key_signup.html})
\end{itemize}

## Tutorials

- [spatial-analysis-and-mapping-workshop.pdf](https://github.com/zalmquist/spatial-analysis-and-mapping-workshop/tree/master/Tutorial)
- [Tutorial_ACS.pdf](https://github.com/zalmquist/spatial-analysis-and-mapping-workshop/tree/master/ACS_Example)
- [WorkBook.pdf](https://github.com/zalmquist/spatial-analysis-and-mapping-workshop/tree/master/WorkBook)


